### 2019-03-31 paper/sex_selection_spacing-ver2.tex
Minor edits

% Demography abstract
\noindent 
Strong son preference is typically associated with shorter birth spacing in the absence of 
sons, but access to sex selection has the potential to reverse this pattern because each 
abortion extends spacing by six to twelve months. 
I introduce an empirical method that simultaneously accounts for how sex selection 
increases both birth spacing and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that birth 
intervals increased substantially over the last four decades, with the most significant
increase among women most likely to use sex selection.
Specifically, well-educated women with no boys now have significantly longer birth
intervals, and more male-biased sex ratios, than similar women with boys. 
Women with no education still follow the standard pattern of short average spacing when 
they have girls and little evidence of sex selection, with medium-educated women showing 
mixed results.
Furthermore, the traditional pattern of longer average spacing with sons appears to arise 
not from fewer very short birth intervals, but rather from the upper end of the distribution.
Finally, northern and western India are responsible for most of the spacing pattern 
reversal, although even the south experienced substantially increased intervals from
sex selection.

@@

Birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination potentially 
changed the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births by six to twelve 
months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, longer birth intervals in the absence of sons, which previously would be taken to 
indicate a lower son preference, now may instead arise because couples with a strong son 
preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case with its long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
Mortality risk is higher for females than males for most ages groups, resulting in
an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Furthermore, fertility has declined substantially, to the point where it is now at, or even 
below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
The combination of falling fertility, strong son preference, and expanding access to 
prenatal sex determination has lead to dramatic increases in the males-to-females ratio at 
birth in India from sex selection over the last three decades
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,Bongaarts2013,Portner2015b,Jayachandran2017}.

@@

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son preference, 
there has so far been no systematic examination of what effect access to sex selection 
have on spacing.
I address three questions to remedy this shortcoming.

@@

Second, have the shapes of the survival curves changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivation behind an increase in average spacing is different if it
comes mainly from parents trying to avoid very short birth intervals, which, as mentioned,
are associated with worse child outcomes, or from continued use of sex selection to
ensure a son.

@@

There are four main results.
First, the intervals between births have increased for all education groups over the four 
decades.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection---well-educated women with no
sons---have seen the largest increases in birth intervals.
As a result, well-educated women now show a complete reversal of the traditional spacing 
pattern, with the longest birth intervals for women with no sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Fourth, the traditional pattern of longer average spacing with sons appear to arise from 
more very long intervals, rather than from fewer very short birth intervals; across 
education groups, there appears to be little difference in the likelihood of having a very 
short birth interval across sex compositions, especially for lower parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with this approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

Average spacing is a convenient way to understand the overall changes in behavior but may 
hide crucial differences in the distribution of spacing.
Specifically, whether increases in average spacing come from fewer births after very short 
intervals, from an even longer spacing of births with already long spacing, or a general 
increase in spacing have important implications for what effects the increases in average 
spacing are likely to have and how to interpret the motivation for the changes.
This section, therefore, examines the distribution of spacing within spells.
I show survival curves conditional on predicted parity progression rather than standard 
survival curves, which allows direct comparisons of the distribution of spacing across 
groups, independent of differences in parity progression probabilities.

@@

What stands out for the ``East'' is the rapid increase in the lengths of birth intervals,
especially for the second spell, where the increases have been more
than a year, independently of whether the first-born was a boy or a girl.
The increases in birth intervals are slightly smaller for the third spell but still 
substantial.
Although the spacing pattern is consistent with some son preference, there is little 
evidence of sex-selective abortions.
Furthermore, despite the ``South'' is generally considered the low fertility region, 
the reduction in fertility has been even faster in the ``East.''
For example, only half of the urban women with a first-born son are now predicted to 
have a second child by the end of the spell.

>>>>

% Demography abstract
\noindent 
Strong son preference is typically associated with shorter birth spacing in the absence of 
sons, but access to sex selection has the potential to reverse this pattern because each 
abortion extends spacing by six to twelve months. 
I introduce an empirical method that simultaneously accounts for how sex selection 
increases both birth spacing and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that birth 
intervals increased substantially over the last four decades, with the most significant
increases among women most likely to use sex selection.
Specifically, well-educated women with no boys now have significantly longer birth
intervals, and more male-biased sex ratios, than similar women with boys. 
Women with no education still follow the standard pattern of short average spacing when 
they have girls and little evidence of sex selection, with medium-educated women showing 
mixed results.
Furthermore, the traditional pattern of longer average spacing with sons appears to arise 
not from fewer very short birth intervals, but rather from the upper end of the distribution.
Finally, northern and western India are responsible for most of the spacing pattern 
reversal, although even the south experienced substantially increased intervals from
sex selection.

@@

Birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination potentially 
changed the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births by six to twelve 
months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, longer birth intervals in the absence of sons, which previously would be taken as 
an indicator of lower son preference, may now instead arise because couples with a strong 
son preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce an empirical method that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case with its long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
Mortality risk is higher for females than males for most ages groups, resulting in
an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Furthermore, fertility has declined substantially, to the point where it is now at, or even 
below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
The combination of falling fertility, strong son preference, and expanding access to 
prenatal sex determination has lead to dramatic increases in the males-to-females ratio at 
birth over the last three decades
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,Bongaarts2013,Portner2015b,Jayachandran2017}.

@@

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son preference, 
there has so far been no systematic examination of what effect access to sex selection 
have on spacing.
I address three questions to remedy this shortcoming in the literature.

@@

Second, have the shapes of the survival curves changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivations that lead to an increase in average spacing are different if 
the increase comes mainly from parents trying to avoid very short birth intervals or from 
continued use of sex selection to ensure a son.

@@

There are four main results.
First, the intervals between births have increased for all education groups over the four 
decades.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection---well-educated women with no
sons---have seen the largest increases in birth intervals.
As a result, well-educated women now show a reversal of the traditional spacing 
pattern, with the longest birth intervals for women with no sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Finally, the traditional pattern of longer average spacing with sons appear to arise from 
more very long intervals, rather than from fewer very short birth intervals; across 
education groups, there appears to be little difference in the likelihood of having a very 
short birth interval across sex compositions, especially for lower parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

Average spacing is a convenient way to understand the overall changes in behavior but may 
hide crucial differences in the distribution of spacing.
Specifically, whether increases in average spacing come from fewer births after very short 
intervals, from an even longer spacing of births with already long spacing, or from a 
general increase in spacing have important implications for what effects the increases in 
average spacing are likely to have and how to interpret the motivation for the changes.
This section, therefore, examines the distribution of spacing within spells.
I show survival curves conditional on predicted parity progression, which allows direct 
comparisons of the distribution of spacing across groups independent of differences in 
parity progression probabilities, rather than standard survival curves.

@@

What stands out for the ``East'' is the rapid increase in the lengths of birth intervals,
especially for the second spell, where the increases have been more
than a year, independently of whether the first-born was a boy or a girl.
The increases in birth intervals are slightly smaller for the third spell but still 
substantial.
Although the spacing pattern is consistent with some son preference, there is little 
evidence of sex-selective abortions.
Furthermore, despite the ``South'' generally considered the low fertility region, 
the reduction in fertility has been even faster in the ``East.''
For example, only half of the urban women with a first-born son are now predicted to 
have a second child by the end of the spell.
